Active Learning for SAT Solver Benchmarking

### 2024-05-24 paper_jar/paper.tex
section 5

In this section, we evaluate our active-learning framework.
First, we analyze and tune the different sub-routines of our framework on the tuning dataset.
Next, we evaluate the best configurations with the full dataset.
Finally, we analyze the importance of different instance families to our framework.

@@

Our experiments follow the evaluation framework introduced in Section~\ref{sec:exdesign:eval}.
Fig.~\ref{fig:e2eallsolvers} shows the performance of the approaches from Section~\ref{sec:exdesign:hyper} on $O_{\operatorname{rt}}$-$O_{\operatorname{acc}}$-diagrams for the hyperparameter-tuning dataset.
Evaluating a particular configuration with Algorithm~\ref{alg:eval} returns a point $\left(O_{\operatorname{rt}},\, O_{\operatorname{acc}}\right)$.
We do not show intermediate results of the active-learning procedure but only the final results after stopping.
The plotted lines represent the best-performing configurations per ranking approach (Fig.~\ref{fig:annitraincolorranking}), selection approach (Fig.~\ref{fig:annitraincolorselection}), and stopping criterion (Fig.~\ref{fig:annitraincolorstopping}).
In particular, we show the Pareto front, i.e., of all configurations that share a particular value of the plotted hyperparameter, we take the maximum ranking accuracy over all remaining hyperparameters \emph{not} displayed in the corresponding plot.

Regarding ranking approaches (Fig.~\ref{fig:annitraincolorranking}), using the predicted $s_3$-induced run\-time-label rank\-ing consistently outperforms the partially observed PAR-2 ranking for each possible value of the trade-off parameter~$\delta$.
This outcome is expected since selection decisions are not random.
For example, we might sample more instances of one family if it benefits discrimination of solvers.
While the partially observed PAR-2 score is skewed, the prediction model can account for this.

Regarding the selection approaches (Fig.~\ref{fig:annitraincolorselection}), uncertainty sampling performs best in most cases.
However, information-gain sampling is beneficial if runtime is strongly favored (small $\delta$; runtime fraction less than \SI{5}{\%}).
This result aligns with our expectations:
Information-gain sampling selects instances that maximize the expected reduction in entropy.
This means we sample instances revealing similarities between solvers rather than differences, which helps to build a confident model quickly.
However, the method cannot select helpful instances for distinguishing solvers later.
Random sampling performs reasonably well but is outperformed by uncertainty sampling in all cases, showing the benefit of actively selecting instances based on a prediction model.

Regarding the stopping criteria (Fig.~\ref{fig:annitraincolorstopping}), the ranking stopping criterion performs most consistently well.
If accuracy is strongly favored (very high $\delta$), the Wilcoxon stopping criterion performs better.
The subset-size stopping criterion performs reasonably well but does not improve beyond a certain accuracy because of sampling a fixed subset of instances.

@@

  \caption{
    Scatter plot comparing different instantiations of trade-off parameter~$\delta$ for our active-learning framework on the hyperparameter-tuning dataset.
    The x-axis shows the fraction of runtime $O_{\operatorname{rt}}$ of the sample, while the y-axes show the fraction of instances sampled and ranking accuracy, respectively.
    The color indicates the weighting between different optimization goals $\delta \in \left[0, 1\right]$.
    The larger $\delta$, the more we favor accuracy over runtime.
  }
  \label{fig:annitrainoptgoal}
\end{figure}

Fig.~\ref{fig:annitrainoptgoalruntime} shows an interesting consequence of weighting our optimization goals:
If we, on the one hand, desire to get a \emph{rough} estimate of a solver's performance fast (low $\delta$), approaches favor selecting many \emph{easy} instances.
In particular, the fraction of sampled instances is larger than the fraction of runtime.
By having many observations, it is easier to build a model.
If we, on the other hand, desire to get a \emph{good} estimate of a solver's performance in a moderate amount of time (high $\delta$), approaches favor selecting few, \emph{difficult} instances.
In particular, the fraction of instances is smaller than the fraction of runtime.

Furthermore, Fig.~\ref{fig:annitrainoptgoalacc} reveals which values make the most sense for $\delta$.
The range $\delta \in \left[0.2, 0.8\right]$, thereby, corresponds to the points with a runtime fraction between \SI{0.03} and \SI{0.22}.
We consider this region to be most promising, analogous to the \emph{elbow} method in cluster analysis~\cite{kodinariya2013review}.

@@

Having selected the most promising hyperparameters, we run our active-learning experiments on the complete Anniversary Track dataset (5355 instances).
The aforementioned range $\delta \in \left[0.2, 0.8\right]$ only results in two distinct configurations.
The best-performing approach for $\delta \in \left[0.2, 0.7\right]$ uses the predicted runtime-label ranking, information-gain sampling, and ranking stopping criterion.
It can predict a new solver's PAR-2 ranking with \SI{90.48}{\%} accuracy ($O_{\operatorname{acc}}$) in only \SI{5.41}{\%} of the full evaluation time ($O_{\operatorname{rt}}$).
The best-performing approach for $\delta \in (0.7, 0.8]$ uses the predicted runtime-label ranking, uncertainty sampling, and ranking stopping criterion.
It can predict a new solver's PAR-2 ranking with \SI{92.33}{\%} accuracy ($O_{\operatorname{acc}}$) in only \SI{10.35}{\%} of the full evaluation time ($O_{\operatorname{rt}}$).

\begin{table}[tp]
  \centering
  \caption{
    Performance comparison (on the full dataset) of the best-performing AL approach (\emph{AL}), random sampling of the same runtime fraction with 1000 repetitions (\emph{Random}), and statically selecting the instances most frequently sampled by active-learning approaches (\emph{Most Frequent})
  }
  \label{tab:fulldataset}
  \begin{tabular}{
    >{\arraybackslash}p{0.4\textwidth}
    S[table-format=2.2,table-column-width=0.15\textwidth]
    S[table-format=2.2,table-column-width=0.15\textwidth]
    S[table-format=2.2,table-column-width=0.15\textwidth]
  }
    \toprule
    {For $\delta \in \left[0.2, 0.7\right]$} & {AL} & {Random} & {Most Frequent} \\
    \midrule
    Sampled Runtime Fraction (\%) & 5.41 & 5.43 & 5.44 \\
    Sampled Instance Fraction (\%) & 26.53 & 5.43 & 27.75 \\
    Ranking Accuracy (\%) & 90.48 & 88.54 & 81.08 \\
    \midrule
    {For $\delta \in (0.7, 0.8]$}  & {AL} & {Random} & {Most Frequent} \\
    \midrule
    Sampled Runtime Fraction (\%) & 10.35 & 10.37 & 10.37 \\
    Sampled Instance Fraction (\%) & 5.24 & 10.37 & 36.96 \\
    Ranking Accuracy (\%) & 92.33 & 91.61 & 84.52 \\
    \bottomrule
  \end{tabular}
\end{table}

Table~\ref{tab:fulldataset} shows how both active-learning approaches (column \emph{AL}) compare against two static baselines:
\emph{Random} samples instances until it reaches roughly the same fraction of runtime as the AL benchmark sets.
We repeat sampling 1000 times and report average results.
\emph{Most Frequent} uses a static benchmark set consisting of those instances most frequently sampled by our active learning approach.
In particular, we consider the average sampling frequency over all solvers and Pareto-optimal active-learning approaches.

Both our AL approaches perform better than random sampling.
However, the performance differences are not significant regarding a Wilcoxon signed-rank test with $\alpha = 0.05$ and also depend on the fraction of sampled runtime (cf. Fig.~\ref{fig:annitraincolorselection}).
A clear advantage of our approach is, though, that it indicates when to stop adding further instances, depending on the trade-off parameter~$\delta$.
While the active-learning results are less strong on the full dataset than on the smaller tuning dataset, they still show the benefit of making benchmark selection dependent on the solvers to distinguish.

A static benchmark using the most frequently AL-sampled instances performs poorly, though, compared to active learning and random sampling.
This outcome is somewhat expected since the static benchmark does not reflect the right balance of instance families:
Families whose instances are uniform-randomly selected by AL, e.g., for different solvers, appear less often in this benchmark than families where some instances are sampled more often than others.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=0.6\textwidth]{../plots/instoccs.pdf}
  \caption{Frequency of how often individual problem instances are selected within all active-learning benchmarks. Instances that are chosen more frequently are intuitively more important to distinguish solver runtimes.}
  \label{fig:instoccs}
\end{figure}

\begin{figure}[tb]
  \centering
  \resizebox{0.85\textwidth}{!}{
    \graphicspath{{../plots/}}
    \input{../plots/anni_final_families.pgf}
  }
  \caption{
    Scatter plot showing the \emph{importance} of different instance families to our framework on the full dataset.
    The x-axis shows the frequency of instance families in the dataset.
    The y-axis shows the average frequency of instance families in the samples selected by active learning.
    The dashed line represents families that occur with the same frequency in the dataset and samples.
  }
  \label{fig:annifinalfamilies}
\end{figure}

We analyze the frequency of individual instances being selected by our active-learning approach.
Fig.~\ref{fig:instoccs} shows the distribution of how often individual instances are selected within all active-learning benchmarks.
We observe that a small fraction of instances is selected very often, while most instances are selected only a few times.
There is no fixed subset of instances that can perfectly distinguish all solvers, which shows the usefulness of employing an active-learning-based strategy.

Selection decisions of our approach also reveal the importance of different instance families to our framework.
Fig.~\ref{fig:annifinalfamilies} shows the occurrence of instance families within the dataset and the benchmarks created by active learning.
We use the best-performing configurations for all $\delta \in \left[0, 1\right]$ and examine the selection decisions by the active-learning approach on the SAT Competition~2022 Anniversary Track dataset~\cite{sat2022}.
While most families appear with the same fraction in the dataset and the sampled benchmarks, a few outliers need further discussion.
Problem instances of the families \emph{fpga}, \emph{quasigroup-completion}, and \emph{planning} are especially helpful to our framework in distinguishing solvers.
Instances of these families are selected over-proportionally in comparison to the full dataset.
In contrast, instances of the largest family, i.e., \emph{hardware-verification}, roughly appear with the same fraction in the dataset and the sampled benchmarks.
Finally, instances of the family \emph{cryptography} are less important in distinguishing solvers than their vast weight in the dataset suggests.
A possible explanation is that these instances are very similar, such that a small fraction of them is sufficient to estimate a solver's performance on all of them.

>>>>

In this section, we evaluate our active learning framework.
First, we analyze and tune the different subroutines of our framework on the tuning dataset.
Next, we evaluate the best configurations on the full dataset.
Finally, we analyze the importance of different instance families for our framework.

@@

Our experiments follow the evaluation framework introduced in Section~\ref{sec:exdesign:eval}.
Fig.~\ref{fig:e2eallsolvers} shows the performance of the approaches from Section~\ref{sec:exdesign:hyper} on $O_{\operatorname{rt}}$-$O_{\operatorname{acc}}$ plots for the hyperparameter tuning dataset.
Evaluating a given configuration with the algorithm~\ref{alg:eval} returns a point $\left(O_{\operatorname{rt}},\, O_{\operatorname{acc}}\right)$.
We do not show intermediate results of the active learning procedure, but only the final results after stopping.
The plotted lines represent the best performing configurations per ranking approach (Fig.~\ref{fig:annitraincolorranking}), selection approach (Fig.~\ref{fig:annitraincolorselection}), and stopping criterion (Fig.~\ref{fig:annitraincolorstopping}).
In particular, we show the Pareto front, i.e., from all configurations that share a particular value of the plotted hyperparameter, we take the maximum ranking accuracy over all remaining hyperparameters \emph{not} shown in the corresponding plot.

Regarding ranking approaches (Fig.~\ref{fig:annitraincolorranking}), using the predicted $s_3$-induced run\-time-label ranking consistently outperforms the partially observed PAR-2 ranking for every possible value of the tradeoff parameter~$\delta$.
This result is expected, since selection decisions are not random.
For example, we might sample more instances of a family if it benefits the discrimination of solvers.
While the partially observed PAR-2 score is biased, the prediction model can account for this.

Regarding the selection approaches (Fig.~\ref{fig:annitraincolorselection}), uncertainty sampling performs best in most cases.
However, information-gain sampling is advantageous when runtime is strongly favored (small $\delta$; runtime fraction less than \SI{5}{\%}).
This result is consistent with our expectations:
Information-gain sampling selects instances that maximize the expected reduction in entropy.
This means that we sample instances that reveal similarities between solvers rather than differences, which helps to build a confident model quickly.
However, the method cannot select instances that are useful for distinguishing solvers later.
Random sampling performs reasonably well, but is outperformed by uncertainty sampling in all cases, demonstrating the benefit of actively selecting instances based on a prediction model.

Regarding the stopping criteria (Fig.~\ref{fig:annitraincolorstopping}), the ranking stopping criterion performs most consistently well.
When accuracy is strongly favored (very high $\delta$), the Wilcoxon stopping criterion performs better.
The subset-size stopping criterion performs reasonably well, but does not improve beyond a certain accuracy because it samples a fixed subset of instances.

@@

  \caption{
    Scatterplot comparing different instantiations of the tradeoff parameter~$\delta$ for our active learning framework on the hyperparameter tuning dataset.
    The x-axis shows the fraction of runtime $O_{\operatorname{rt}}$ of the sample, while the y-axes show the fraction of sampled instances and the ranking accuracy, respectively.
    The color indicates the weighting between different optimization objectives $\delta \in \left[0, 1\right]$.
    The larger $\delta$, the more we favor accuracy over runtime.
  }
  \label{fig:annitrainoptgoal}
\end{figure}

Fig.~\ref{fig:annitrainoptgoalruntime} shows an interesting consequence of weighting our optimization goals:
On the one hand, if we want to get a \emph{rough} estimate of a solver's performance quickly (low~$\delta$), the approaches favor selecting many \emph{easy} instances.
In particular, the fraction of sampled instances is larger than the fraction of runtime.
By having many observations, it is easier to build a model.
On the other hand, if we want to get a \emph{good} estimate of a solver's performance in a moderate amount of time (high~$\delta$), approaches favor selecting few, \emph{difficult} instances.
In particular, the fraction of instances is smaller than the fraction of runtime.
Furthermore, Fig.~\ref{fig:annitrainoptgoalacc} shows which values of $\delta$ make the most sense.
The range $\delta \in \left[0.2, 0.8\right]$ corresponds to the points with a runtime fraction between \SI{0.03} and \SI{0.22}.
We consider this region to be the most promising, analogous to the \emph{elbow} method in cluster analysis~\cite{kodinariya2013review}.

@@

After selecting the most promising hyperparameters, we run our active learning experiments on the complete Anniversary Track dataset (5355 instances).
The above range $\delta \in \left[0.2, 0.8\right]$ results in only two different configurations.
The best performing approach for $\delta \in \left[0.2, 0.7\right]$ uses the predicted runtime label ranking, information-gain sampling, and ranking stopping criterion.
It can predict the PAR-2 ranking of a new solver with \SI{90.48}{\%} accuracy ($O_{\operatorname{acc}}$) in only \SI{5.41}{\%} of the total evaluation time ($O_{\operatorname{rt}}$).
The best performing approach for $\delta \in (0.7, 0.8)$ uses the predicted runtime label ranking, uncertainty sampling, and ranking stopping criterion.
It can predict the PAR-2 ranking of a new solver with \SI{92.33}{\%} accuracy~($O_{\operatorname{acc}}$) in only \SI{10.35}{\%} of the total evaluation time~($O_{\operatorname{rt}}$).

\begin{table}[tp]
  \centering
  \caption{
    Performance comparison (on the full dataset) of the best performing AL approach (\emph{AL}), random sampling of the same runtime fraction with 1000 repetitions (\emph{Random}), and static selection of the instances most frequently sampled by active learning approaches (\emph{Most Frequent}).
  }
  \label{tab:fulldataset}
  \begin{tabular}{
    >{\arraybackslash}p{0.4\textwidth}
    S[table-format=2.2,table-column-width=0.15\textwidth]
    S[table-format=2.2,table-column-width=0.15\textwidth]
    S[table-format=2.2,table-column-width=0.15\textwidth]
  }
    \toprule
    {For $\delta \in \left[0.2, 0.7\right]$} & {AL} & {Random} & {Most Frequent} \\
    \midrule
    Sampled Runtime Fraction (\%) & 5.41 & 5.43 & 5.44 \\
    Sampled Instance Fraction (\%) & 26.53 & 5.43 & 27.75 \\
    Ranking Accuracy (\%) & 90.48 & 88.54 & 81.08 \\
    \midrule
    {For $\delta \in (0.7, 0.8]$}  & {AL} & {Random} & {Most Frequent} \\
    \midrule
    Sampled Runtime Fraction (\%) & 10.35 & 10.37 & 10.37 \\
    Sampled Instance Fraction (\%) & 5.24 & 10.37 & 36.96 \\
    Ranking Accuracy (\%) & 92.33 & 91.61 & 84.52 \\
    \bottomrule
  \end{tabular}
\end{table}

Table~\ref{tab:fulldataset} shows how the two active learning approaches (column \emph{AL}) compare to two static baselines:
\emph{Random} samples instances until it reaches roughly the same fraction of runtime as the AL benchmark sets.
We repeat the sampling 1000 times and report the average results.
\emph{Most Frequent} uses a static benchmark set consisting of the instances most frequently sampled by our active learning approach.
In particular, we consider the average sampling frequency across all solvers and Pareto-optimal active learning approaches.

Both of our AL approaches outperform random sampling.
However, the performance differences are not significant with respect to a Wilcoxon signed-rank test with $\alpha = 0.05$, and also depend on the fraction of the runtime sampled (see Fig.~\ref{fig:annitraincolorselection}).
However, a clear advantage of our approach is that it indicates when to stop adding new instances, depending on the tradeoff parameter~$\delta$.
While the active learning results on the full dataset are not as strong as on the smaller tuning dataset, they still show the benefit of making the benchmark selection dependent on the solvers to be distinguished.

A static benchmark utilizing the most frequently AL-sampled instances exhibits suboptimal performance when compared to active learning and random sampling.
This outcome is somewhat anticipated, given that the static benchmark fails to reflect an appropriate balance of instance families.
For instance, families whose instances are uniformly randomly selected by AL, e.g., for different solvers, appear less frequently in this benchmark than families where some instances are sampled more often than others.
% MI: What? I don't understand this sentence.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=0.6\textwidth]{../plots/instoccs.pdf}
  \caption{The frequency with which individual problem instances are selected within all active-learning benchmarks provides insight into the relative importance of these instances. Instances that are chosen more frequently are intuitively more important for distinguishing solver runtimes.}
  \label{fig:instoccs}
\end{figure}

\begin{figure}[tb]
  \centering
  \resizebox{0.85\textwidth}{!}{
    \graphicspath{{../plots/}}
    \input{../plots/anni_final_families.pgf}
  }
  \caption{
    The scatter plot illustrates the relative importance of different instance families for ranking. The x-axis depicts the frequency of instance families in the dataset, while the y-axis represents the average frequency of instance families in the samples selected by active learning. The dashed line represents families that occur with the same frequency in the dataset and samples.
  }
  \label{fig:annifinalfamilies}
\end{figure}

Figure~\ref{fig:instoccs} illustrates the distribution of the frequency with which individual instances are selected across all active learning benchmarks.
It can be observed that a small proportion of instances is selected with great frequency, while the majority of instances are selected on only a few occasions.
This indicates that there is no fixed subset of instances that can perfectly distinguish all solvers, which highlights the value of employing an active learning-based strategy.

The selection decisions of our approach also reveal the importance of different instance families to our framework.
Fig.~\ref{fig:annifinalfamilies} illustrates the occurrence of instance families within the dataset and the benchmarks created by active learning.
We utilize the most effective configurations for all values of $\delta \in \left[0, 1\right]$ and examine the selection decisions made by the active learning approach on the SAT Competition~2022 Anniversary Track dataset~\cite{sat2022}.
While the majority of families exhibit the same fraction in the dataset and the sampled benchmarks, a few outliers require further discussion.
Instances of the families \emph{fpga}, \emph{quasigroup-completion}, and \emph{planning} are particularly useful for distinguishing solvers within our framework.
These instances are selected in greater proportion than in the full dataset.
In contrast, instances of the largest family, i.e., \emph{hardware-verification}, appear with approximately the same fraction in the dataset and the sampled benchmarks.
Finally, instances of the family \emph{cryptography} are less important in distinguishing solvers than their substantial representation in the dataset would suggest.
One possible explanation for this discrepancy is that these instances are highly similar, such that a small fraction of them is sufficient to estimate a solver's performance on all of them.
